---
title: "Assignment 1 of Reproducible Research"
author: "Alex Ge, alexgecontrol@qq.com"
date: "Tuesday, January 13, 2015"
output: html_document
---

## Overview

In this assignment, three main problems--effects of imputing on mean and medians of total number of steps taken per day, average daily activity pattern and activity patterns between weekdays and weekends--are analyzed and reported.

## 0. Set Up R Session

In this assignment I use package *data.table* for data mungling and *ggplot2* for data visualization.
Below is my code for session setup.

```{r SetUpSession, echo=TRUE,warning=FALSE}
# Set working directory:
setwd( "C:/Users/thinkpad/Desktop/00-Mathematics/03-Statistics/01-Basic Statistics/Projects/01-Data-Specialization-JHU-Coursera/04-Reproducible-Research/01-Assignment-One" )
# Package for graphing:
library( ggplot2 )
# Package for data mungling:
library( data.table )
```

## 1. Data Mungling

I created three new datasets from the raw data, namely,
the merged dataset of total number of steps taken per day made from raw and imputed data,
the average steps per interval data and
the average steps per interval data faceted by weekdays and weekends.
Below are the detailed mungling processes for each dataset.

### 1.1. Merged Dataset of Total Number of Steps Taken Per Day Made From Raw and Imputed Data

The mungling process for this dataset is as follows:

```{r DataMunglingOne, echo=TRUE}
# Raw data:
raw_data <- read.csv( "activity.csv" , header = TRUE , stringsAsFactors = FALSE )
raw_data_table <- data.table( raw_data )
# Imputed data:
imputed_data <- raw_data
imputed_data$steps[ is.na( imputed_data$steps ) ] <- median( imputed_data$steps , na.rm = TRUE )
imputed_data_table <- data.table( imputed_data )
# Total steps by day:
total_steps_by_day_raw <- raw_data_table[ , sum( steps ), by = date ]
total_steps_by_day_raw$Imputed <- "False"
total_steps_by_day_imputed <- imputed_data_table[ , sum( steps ), by = date ]
total_steps_by_day_imputed$Imputed <- "True"
total_steps_by_day <- rbind( total_steps_by_day_imputed , total_steps_by_day_raw )
setnames( total_steps_by_day , "V1" , "total.steps" )
total_steps_by_day$Imputed = factor( total_steps_by_day$Imputed , levels = c( "True" , "False" ) )
```

Here the factor attribute *Imputed* is added to facilitate further graphing.

### 1.2. Average Steps Per Interval Data

The mungling process for this dataset is as follows:

```{r DataMunglingTwo, echo=TRUE}
# Average steps by interval:
mean_steps_by_interval <- imputed_data_table[ , mean( steps ), by = interval ]
setnames( mean_steps_by_interval , "V1" , "average.steps" )
```

### 1.3. Average Steps Per Interval Data Faceted by Weekdays and Weekends

The mungling process for this dataset is as follows:

```{r DataMunglingThree, echo=TRUE}
# Average steps by interval faceted by day type:
date <- strptime( imputed_data$date , "%Y-%m-%d" )
days <- weekdays( date , abbreviate = TRUE )
is_weekend <- days %in% c( "ÖÜÁù" , "ÖÜÈÕ" )
mean_steps_by_weekdays <- imputed_data_table[ !is_weekend , mean( steps ), by = interval ]
mean_steps_by_weekdays$Day <- "Weekdays"
mean_steps_by_weekend <- imputed_data_table[ is_weekend , mean( steps ), by = interval ]
mean_steps_by_weekend$Day <- "Weekend"
mean_steps_by_day <- rbind( mean_steps_by_weekdays , mean_steps_by_weekend )
setnames( mean_steps_by_day , "V1" , "average.steps" )
mean_steps_by_day$Day = factor( mean_steps_by_day$Day , levels = c( "Weekdays" , "Weekend" ) )
```

## 2. Mean and Median of Total Steps Per Day Using Raw Data

### 2.1. Histogram

Here is the histogram of total steps per day generated using raw data:

```{r HistogramOfRawData,echo=FALSE,warning=FALSE,message=FALSE,out.width='640px',out.height='360px',fig.align='center' }
# Histogram:
qplot( total.steps , data = total_steps_by_day[ "False" == total_steps_by_day$Imputed , ] ,
       geom = "histogram" , 
       xlab = "Total Steps" , ylab = "Frequency" , main = "Histogram for Total Steps Per Day Using Raw Data" )
```

The code for plot generation is as follows:

```{r HistogramOfRawDataCode,eval=FALSE}
# Histogram:
qplot( total.steps , data = total_steps_by_day[ "False" == total_steps_by_day$Imputed , ] ,
       geom = "histogram" , 
       xlab = "Total Steps" , ylab = "Frequency" , main = "Histogram for Total Steps Per Day Using Raw Data" )
```

### 2.2. Mean and Median

The two statistics are summarized in the following table.

Statistics  |Mean                               |Median
:----------:|:---------------------------------:|:----------------------------------:
Value       |`r mean(total_steps_by_day_raw$V1,na.rm=TRUE)`|`r median(total_steps_by_day_raw$V1,na.rm=TRUE)`

## 3. Average Steps Per Interval

### 3.1. Time Series Visualization

Here is the visualization of the data:

```{r TimeSeries,echo=FALSE,warning=FALSE,message=FALSE,out.width='640px',out.height='360px',fig.align='center' }
# Time series:
qplot( interval , average.steps , data = mean_steps_by_interval , colour = "red" ,
       geom = "line" ,
       xlab = "Interval" , ylab = "Average Steps" , main = "Average Steps by Time Intervals within A Day" ) + 
       guides( colour = FALSE )
```

The code for plot generation is as follows:

```{r TimeSeriesCode,eval=FALSE}
# Time series:
qplot( interval , average.steps , data = mean_steps_by_interval , colour = "red" ,
       geom = "line" ,
       xlab = "Interval" , ylab = "Average Steps" , main = "Average Steps by Time Intervals within A Day" ) + 
       guides( colour = FALSE )
```

### 3.2. The Interval with Maximum Number of Average Steps

The interval with maximum number of average steps is `r mean_steps_by_interval$interval[ mean_steps_by_interval$average.steps == max( mean_steps_by_interval$average.steps ) ]`.

## 4.Imputing Missing Values

### 4.1. The Total Number of Missing Values

The total number of missing values is `r sum( as.integer( is.na( raw_data$steps ) ) )`.

### 4.2. Strategy for Imputing

According to the results of Analysis 2, the data is right skewed. So I choose to impute the data with median.

### 4.3. Imputed Dataset

The raw data is imputed and the new dataset is merged with the original to form the new dataset *total_steps_by_day* as it does in Section 1.1.

### 4.4. Effects of Imputing

**First** let's look at the two histograms of the raw and imputed data:

```{r ComparisonOfHistograms,echo=FALSE,warning=FALSE,message=FALSE,out.width='640px',out.height='360px',fig.align='center' }
# Comparison of histograms:
qplot( total.steps , data = total_steps_by_day , 
       geom = "histogram" , fill = Imputed , position = "dodge" ,
       xlab = "Total Steps" , ylab = "Frequency" , main = "Histogram for Raw/Imputed \'Total Steps by Day\' Data" )
```

The code for plot generation is as follows:

```{r ComparisonOfHistogramsCode,eval=FALSE}
# Comparison of histograms:
qplot( total.steps , data = total_steps_by_day , 
       geom = "histogram" , fill = Imputed , position = "dodge" ,
       xlab = "Total Steps" , ylab = "Frequency" , main = "Histogram for Raw/Imputed \'Total Steps by Day\' Data" )
```

From the above two histograms we can see that my strategy of imputing introduces extra zero observations into the dataset.

**Then** let's compare the statistics used of:

Data    |Mean                                          |Median
:------:|:--------------------------------------------:|:----------------------------------:
Raw     |`r mean(total_steps_by_day_raw$V1,na.rm=TRUE)`|`r median(total_steps_by_day_raw$V1,na.rm=TRUE)`
Imputed |`r mean(total_steps_by_day_imputed$V1)`       |`r median(total_steps_by_day_imputed$V1)`

We know from the above table that **imputing has lower the values of mean and median**.

## 5.Activity Pattern Differences between Weekdays and Weekends

### 5.1. Factor Attribute for Weekdays and Weekends

The factor attribute for weekdays and weekends has been created as it does in Section 1.3.

### 5.2. Average Steps Per Interval Faceted by Weekdays and Weekends

Here is the result of faceting:

```{r ComparisonOfTimeSeries,echo=FALSE,warning=FALSE,message=FALSE,out.width='640px',out.height='360px',fig.align='center' }
# Comparison of time series:
qplot( interval , average.steps , data = mean_steps_by_day , colour = "red" , facets = Day ~ . ,
       geom = "line" ,
       xlab = "Interval" , ylab = "Average Steps" , main = "Average Steps by Time Intervals Faceted by Day" ) + 
       guides( colour = FALSE )
```

The code for plot generation is as follows:

```{r ComparisonOfTimeSeriesCode,eval=FALSE}
# Comparison of time series:
qplot( interval , average.steps , data = mean_steps_by_day , colour = "red" , facets = Day ~ . ,
       geom = "line" ,
       xlab = "Interval" , ylab = "Average Steps" , main = "Average Steps by Time Intervals Faceted by Day" ) + 
       guides( colour = FALSE )
```
